---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
    number_sections: yes
    toc: yes
    toc_depth: 3

---


# Loading and preprocessing the data
## Load the data, checks that the activity.csv file is present.

```{r getdata}
if (file.exists("activity.zip"))
{ 
	activity_data <- read.csv(unz("activity.zip", "activity.csv"))
} else 
{
    return(message("Activity data files (activity.zip) missing"))
}
```

### Quick review of the data
```{r reviewdata}
summary(activity_data)
str(activity_data)
```

### Aggregate and Plot total steps by day
```{r aggregate_and_plot_data}
activity_steps_totals <- aggregate(steps ~ date, data=activity_data, FUN = function(x) c(totals=sum(x))) 
hist(activity_steps_totals$steps, main="Histogram for Activity - Total Steps by Day", xlab="Number of Steps", ylab="Number of Days", breaks="freedman-diaconis", col="red")
```

# What is mean total number of steps taken per day?

##Mean number of Steps per Day: 
```{r mean_steps}
mean(activity_steps_totals$steps)
```

##Median number of Steps per Day: 
```{r median_steps}
median(activity_steps_totals$steps)
```


# What is the average daily activity pattern?

## Aggregate and Plot steps by interval
```{r plot_activity}
activity_steps_averages <- aggregate(steps ~ interval, data=activity_data, FUN = function(x) c(average=mean(x))) 
plot(activity_steps_averages$steps, type="l", col="blue", ylab="Number of Steps", xlab="Interval Number", main="Line Plot of Steps by 5 minute Intervals")
```

## Interval with greatest average number of steps
```{r interval_highmark}
activity_steps_averages[which(activity_steps_averages$steps == max(activity_steps_averages$steps)),]
```


# Imputing missing values
##Count NAs: 
```{r count_na}
sum(is.na(activity_data$steps))
```

## Repopulate NA's with arbitrary values...

Simplest approach is to use the average value from each timestep, as we have already calculated that value
```{r replace_na_with_imputed_values}
activity_steps <- data.frame(activity_data$steps)
activity_steps[is.na(activity_steps),] <- ceiling(tapply(X=activity_data$steps,FUN=mean,INDEX=activity_data$interval, na.rm=TRUE))
activity_steps_imputed <- cbind(activity_steps, activity_data)
colnames(activity_steps_imputed) <- c("Steps_Imputed","Steps_Original", "Date", "Interval")

activity_steps_imputed_totals <- aggregate(Steps_Imputed ~ Date, data=activity_steps_imputed, FUN = function(x) c(totals=sum(x))) 
```

## Plot Histogram with Imputed values for NA's
```{r plot_with_imputed_data}
hist(activity_steps_imputed_totals$Steps_Imputed, main="Histogram for Activity - Total Steps (with Imputed Values)", 	xlab="Number of Steps", 	ylab="Number of Days", 	breaks="freedman-diaconis",	col="pink")
```	

##Mean number of Steps per Day based on Imputed Data: 
```{r mean_with_imputed_data}
mean(activity_steps_imputed_totals$Steps_Imputed)
```

##Median number of Steps per Day based on Imputed Data:
```{r median_with_imputed_data}
median(activity_steps_imputed_totals$Steps_Imputed)
```

Percent change in Mean number of steps per day: 
```{r percent_change_means}
(mean(activity_steps_imputed_totals$Steps_Imputed)-mean(activity_steps_totals$steps))/mean(activity_steps_totals$steps)*100
```

Percent change in Median number of steps per day: 
```{r percent_change_medians}
(median(activity_steps_imputed_totals$Steps_Imputed)-median(activity_steps_totals$steps))/median(activity_steps_totals$steps)*100
```

We see a small increase in the mean and median step count when we impute the average in place of the NA values.

## Are there differences in activity patterns between weekdays and weekends?

Differentiate weekdays (0)  vs weekends (1) in a 'day' field
```{r separate_weekdays_and_weekends}
tm1.lt <- as.POSIXlt(activity_data$date)
activity_data$day <- tm1.lt$wday
activity_data$day[activity_data$day <= 5] <- "0" # Day = 0 if Weekday
activity_data$day[activity_data$day > 5] <- "1"  # Day = 1 if Weekend
activity_steps_averages_by_weekday <- aggregate( data=activity_data, steps ~ day + interval, FUN=mean )
activity_data_weekdays <- activity_steps_averages_by_weekday[activity_steps_averages_by_weekday$day==0,]
activity_data_weekends <- activity_steps_averages_by_weekday[activity_steps_averages_by_weekday$day==1,]
```

## Plot Weekday and Weekend Interval Activity Historgrams
```{r plot_weekdays_vs_weekends, fig.width = 10, fig.height = 10}
old.par <- par(mfrow=c(2, 1))
plot(activity_data_weekends$steps,type="l", col="blue", ylab="Number of Steps", xlab="Interval Number", main="Line Plot of Steps by 5 minute Intervals - Weekends")
plot(activity_data_weekdays$steps,type="l", col="blue", ylab="Number of Steps", xlab="Interval Number", main="Line Plot of Steps by 5 minute Intervals - Weekdays")
```

Generally, the data suggests that people are more active on the weekends.  

There is more early morning activity and the activity is more constant (reduced spikes/troughs) over the weekends when compared to the weekday chart.
